An Interactive Tutorial on Hash Array Mapped Tries

### 2024-08-23 paper/paper.tex
remove optimized kw maps

\authorinfo{ Ambrose Bonnaire-Sergeant}
           { Indiana University Bloomington }
           { abonnair@indiana.edu}

@@

\title{Towards Optimized Keyword Hash Maps for Clojure}

@@

\begin{abstract}
Clojure provides a suite of persistent data structures
implemented by Hickey based on previous work by Bagwell.
There are several implementations of persistent hash maps
included in Clojure and provided by open source source libraries
(like data.int-map), however there is no specific
implementation for the most common case of hash map:
relatively small hash maps (less than 32 entries)
consisting of keyword keys.
Keywords are effectively interned strings that are designed
to be keys in maps.
This paper presents how the current implementations
of hash maps handle keyword usage, 
and present several experiments of specialized hash maps
handling just keyword keys.

\end{abstract}

@@

\paragraph{Contributions}
This paper offers several contributions.

\begin{enumerate}
  \item We give an approachable walkthrough of mechanics behinds
      HAMTs.
  \item We describe the internals of Clojure's persistent HAMT implementation,
    via our pure-Clojure reimplementation.
  \item We present a port of Clojure's HAMT from Java to Clojure
    for pedagogical purposes,
    featuring trie visualization (Figure~\label{port-visualize}).
  \item We evaluate a Clojure-specific optimization for Clojure's HAMT.
  %\item We speculate on future directions for HAMT's.
\end{enumerate}

@@

@@

Search is straightforward, and works as described
in Section~\ref{walkthrough}.

@@

\section{Experiments}
\label{experiments}

We prototyped an optimization based on
\empty{just-in-time compilation} techniques.
Leveraging Clojure's \texttt{case} statement,
we created specialized hash array mapped tries
for specific keysets.
We decide which keysets based on runtime frequency.

Maps with keyword keys are very common in Clojure.
A type system for Clojure has special support
for such maps~\cite{bonnaire2016practical}
and the \texttt{clojure.spec} library\footnote{\url{http://clojure.org/guides/spec}}
has exposes special primitives to generate and verify
such maps.

%\begin{figure}
%\begin{verbatim}
%(assoc [this key val]
%  (cond
%    (keyword? key)
%    (case key
%      ...   ;; specific branches for each key
%      ;; default case
%      )
%
%    :else
%\end{verbatim}
%\caption{Pseudocode for \texttt{assoc} in optimized maps}
%\end{figure}
%
%\begin{figure}
%\begin{verbatim}
%(get [this key]
%  (cond
%    (keyword? key)
%    (case key
%      ...   ;; specific branches for each key
%      ;; default case
%      )
%
%    :else
%\end{verbatim}
%\caption{Pseudocode for \texttt{get} in optimized maps}
%\end{figure}

\begin{figure*}
\includegraphics[width=18cm]{big-tree}
\caption{
Trie visualization in Clojure port, for map
\texttt{\{:a 1 :b 2 :c 3 :d 4 :e 5 :f 6 :g 7 :h 8 :i 9 :j 10\}}
(See Appendix~\ref{hash-examples} for corresponding hashes).
}
\label{big-trie-vis}
\end{figure*}

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}[
      legend pos=north west,
      xlabel=Number of Extra keys,
      ylabel=Time (msec)]
      \addplot+[sharp plot] coordinates
      {(0,837.900884) (100,1435.768863) (200,1447.167303) (300,1505.881845) (400,1331.124945) 
        (500,1759.42348 8) (600,1387.997231) (700,1729.756068) (800,1784.948714) (900,1481.740526) 
        (1000,1521.329677) (1100,1855.281003) (1200,2573.8774) (1300,2049.42444) (1400,1817.21074) 
        (1500,2023.47916) (1600,1936.875666) (1700,2129.781288) (1800,1757.819692) (1900,2973.424349)};
      \addlegendentry{Plain map}
      \addplot+[sharp plot] coordinates
{
(0,840.862397) (100,894.77844) (200,921.830137) (300,904.195596) (400,868.788413) 
(500,849.688625) (600,857.498675) (700,884.393927) (800,1099.811693) (900,936.40417) 
(1000,1079.314484) (1100,1598.098832) (1200,1334.392182) (1300,1212.918047) (1400,1047.268085) 
(1500,1110.733413) (1600,1232.646281) (1700,1181.103518) (1800,1400.642201) (1900,2108.923538)
};
      \addlegendentry{Record}
      \addplot+[sharp plot] coordinates
{
(0,892.363186) (100,972.581187) (200,921.286823) (300,944.139623) (400,850.684909) 
(500,907.794038) (600,835.013332) (700,841.138563) (800,938.415172) (900,960.614487)
 (1000,904.91121) (1100,1569.635548) (1200,1294.22056) (1300,1047.200465) (1400,1112.890638)
 (1500,1203.444674) (1600,1261.062708) (1700,2109.647289) (1800,3016.390694) (1900,3844.42986)
};
      \addlegendentry{Optimized map}
    \end{axis}
  \end{tikzpicture}
  \caption{Running time for keyword keys benchmark,
  varying the number of extra entries in the map.}
  \label{bench-plot}
\end{figure}

\subsection{Approach}

To prototype this approach, we repurposed Clojure's
\texttt{defrecord} construct, which creates maps
specialized maps on a known keyset.
Instead of statically compiling only known keysets,
our approach effectively compiles records at runtime
based the most frequent keyword keysets for that
particular run.

\subsection{Evaluation}

To evaluate our approach, we developed a prototype
that is a simple wrapper around the existing \texttt{PersistentHashMap}
class that intercepts operations on keyword keys.

To minimize the overhead of compiling new classes, we
delegate compilation to a separate thread of execution,
using shared data to communicate which keysets to 
specialize.

We tested three kinds of maps in our benchmark.

\paragraph{Plain map}
This is the Clojure implementation of hash array mapped trie
described in Section~\ref{clojure-hamt}.

\paragraph{Record}
This is a map constructed with Clojure's \texttt{defrecord}
form. It is specialized at compile time for a specific
known set of keyword entries.

\paragraph{Optimized map}
This is our prototype implementation that specializes
maps at runtime.

\begin{figure}
\begin{verbatim}
(defn exercise-bench [f n]
  (loop [i 20
         m (f (into {:a 1 :b 2 :c 3 :d 4
                     :e 5 :f 6 :g 7 :h 8}
                    (map #(vector % %) (range n))))]
    (when-not (zero? i)
      (dotimes [_ 100000] 
        (+ (:a m) (:b m) (:c m) (:d m)
           (:e m) (:f m) (:g m) (:h m)))
      (recur (dec i) (update m :a inc)))))
\end{verbatim}
\caption{Benchmark code that takes a function $f$
that returns the kind of map we are currently
benchmarking (eg., plain, record, or optimized),
and a number $n$, the number of extra keys to fill
the map with.
}
\label{benchmark-code}
\end{figure}

Figure~\ref{benchmark-code} contains the benchmark
code. The purpose of the benchmark is to simulate
a set of map associations followed by many lookups,
since our optimization only triggers after an association
operation.
In this version of the benchmark, we perform 20
associations, each with 400,000 lookups
on keyword entries.
Each iteration has a set number of keyword entries,
and we vary the

\subsection{Results}

Figure~\ref{bench-plot} plots the results.
The \emph{Record} is hand-optimized for the particular
keyword entries in the benchmark, and outperforms
the \emph{Plain map} as expected.
Our \emph{Optimized map} compiles its own specialized
map for the set of keywords, and consistently
beats \emph{Plain map} for 1,500 extra keys and under.
It also stays competitive with the hand-tuned \emph{Record}
for 1,500 and under.

After 1,500 keys, there is an unknown problem with our optimized
implementation. Our hand-written benchmarking scheme
does not compensate for garbage collection and is rather crude,
so we would like to revisit this benchmark in the future to 
investigate this issue.

\subsection{Interpretation}

Our optimization stays mostly competitive with hand-tuned
records. However, there are no wins in our predicted
primary use-case: small maps with only keyword keys.
Our benchmark tested an unusual case: keyword lookups on a
hybrid keyword and integer keyed map.
We constructed it like this to ensure our implementation beat
plain maps in expected cases.

We hope our approach can be improved to cater to our
original vision of speeding up small keyword maps---our
initial experiments show no speedup in this case.

%\begin{verbatim}
%- try associating a number with keywords
%- try generating defrecord classes for keyword
%  maps
%- try splitting keywords and other data into two
%  trees
%- combine RB trees/something else with other maps?
%  - then branch on every `get` if something is a keyword,
%    use the tree optimised for keywords, otherwise use the
%    normal thing.
%- try compiling a `case` statement for all the
%  keywords
%  - try doing asynchronously
%    - then set mutable field to utilize the thunk
%    - while it's compiling, use a PHM in the mean time
%- keep a set of the most common clusters of keyword parameters
%  - a separate thread compiles defrecords for them
%  - then when these maps are created/assoc'd/dissoc'd,
%    they can directly use the defrecords
%\end{verbatim}

@@

Our approach to speeding up the hash-map implementation
is related to \textit{storage strategies}
by Bolz et. al~\cite{Bolz13storagestrategies}.
They observe that dynamically typed languages often use
heterogeneous collections as homogeneous, and present
optimizations to take advantage of this fact.

@@

\section{Future directions}

In future work, we plan to devise several benchmarks
that test real-world usages of small keyword maps.
We will devise several more optimization strategies
based on the optimization described in Section~\ref{experiments},
and evaluate them using these new benchmarks.

Furthermore, we plan to override Clojure's default hash map implementation
in a custom version of Clojure, and run benchmarks
over entire programs.
We plan to find programs that extensively use the ``tagged map''
idiom of using plain maps as records---we conjecture these
are more likely to see speedups if our optimizations
satisfy the micro-benchmarks.

\section{Conclusion}

%\begin{verbatim}
%- summarise the problem
%- recap the history of Clojure's decisions on data structures
%- summarise the results of our experiments.
%- recommendations
%\end{verbatim}

Clojure has a general purpose hash-map implementation, but
programmers are encouraged to hash-maps primarily as records,
which are typically small maps with less than 32 entries
with only keyword keys.

In this paper, we prototyped an optimization that caters
to this usecase, by compiling specialized hash-maps
for specific keyword keysets discovered at runtime.
Our preliminary investigation found we could duplicate
performance for hand-tuned records with hybrid keyword and
integer keyed maps.
In the future, we plan to investigate optimizations
and benchmarks that better target small keyword maps.

\section{Dedication}
This paper is dedicated to the memory of Phil Bagwell.
Thank you for sharing your gifts.

>>>>

\authorinfo{ Ambrose Bonnaire-Sergeant}
           { ambrose@ambrosebs.com }

@@

\title{An Interactive Tutorial on Hash Array Mapped Tries}

@@

\begin{abstract}
Clojure provides a suite of persistent data structures
implemented by Hickey based on previous work by Bagwell.
In this tutorial, we use a ported implementation of
Hickey's Java implementation to Clojure to learn
how Hash Array Mapped Tries work.
\end{abstract}

@@

\paragraph{Contributions}

\begin{enumerate}
  \item We walkthrough the mechanics behind HAMTs.
  \item We describe the internals of Clojure's persistent HAMT implementation.
  \item We present a port of Clojure's HAMT from Java to Clojure
    for pedagogical purposes,
    featuring trie visualization (Figure~\label{port-visualize}).
  %\item We evaluate a Clojure-specific optimization for Clojure's HAMT.
  %\item We speculate on future directions for HAMT's.
\end{enumerate}

@@

\begin{figure*}
\includegraphics[width=18cm]{big-tree}
\caption{
Visualization of
\texttt{\{:a 1 :b 2 :c 3 :d 4 :e 5 :f 6 :g 7 :h 8 :i 9 :j 10\}},
hashes in Appendix~\ref{hash-examples}.
}
\label{big-trie-vis}
\end{figure*}

@@

Search is straightforward and similar to Section~\ref{walkthrough}.

@@

@@

@@

\section{Dedication}
This paper is dedicated to the memory of Phil Bagwell.
